---
title: "R Notebook"
output: 
  html_notebook: 
    fig_height: 7
  pdf_document: 
    fig_height: 7
---

Introduction

Study the power spectra of noise modulated by something periodic.

SETUP
```{r}
rm(list=ls())
setwd("~/WORKSHOP/QAANAAQ/")
```



Define function to make power spectra

```{r}
pth_spec <- function(data)
{
  time <- data[, 1]
  signal <- data[, 2]
  
# Calculate the power spectrum using the FFT
fft_result <- abs(fft(signal)) ^ 2 / length(signal)
half_length <- length(fft_result) %/% 2  # Get the index for the first half
frequencies <- seq(0, 1 / (2 * diff(time)[1]), length.out = half_length)

# Plot the first half of the power spectrum
plot(1/frequencies, fft_result[1:half_length], type = "l", 
     xlab = "1/Frequency", ylab = "Power Spectrum", 
     main = "Power Spectrum of Time Series (First Half)",
     log="xy", pch=19, cex=0.2)
  
  
  return(list("period" = 1/frequencies, "power" = fft_result))
}
```


A test example
```{r}
par(mfrow=c(2,1))
n <- 10000
x= 1:n
period = 500
y <- rnorm(n,0,1)
plot(x,y,pch=19, cex=0.2)
res <- pth_spec(cbind(x,y))
y <- y * (1+cos(2*pi/period*x))
plot(x,y,pch=19, cex=0.2)
#
res <- pth_spec(cbind(x,y))
```


Discussion of results

Since the white noise is ... noise ... there are no dominant frequencies, hence no 'carrier frequency' f_c and so the act of modulating the WN signal with a period function (f_m) should produce side-bands at f_c +/- f_m - that is 'everywhere' - leading to a spectrum with no features, probably just like the original series spectrum without modulation.